\documentclass[a4paper,UKenglish,cleveref, autoref, thm-restate]{lipics-v2021}

\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{refcount} % ラベルの番号を取得するために使用
\newcounter{tempTheoremCounter} % カウンタを一時的に保存するための新しいカウンタを定義

\usepackage{tikz}
\usetikzlibrary{tikzmark, arrows.meta, bending}

\usepackage{CJKutf8} % 日本語を書きたいので
\usepackage{ascmac} % なんか枠を作る
\usepackage{stmaryrd}

\newcommand*{\Ja}[1]{%
  \begin{CJK}{UTF8}{ipxm}#1\end{CJK}%
}

\newenvironment{Jcomment}%
{\begin{screen}\begin{CJK}{UTF8}{ipxm}}%
{\end{CJK}\end{screen}}

\usepackage[most]{tcolorbox} % for \begin{tcolorbox}\end{tcolorbox}
% \usepackage[bibliography=common]{apxproof}
\usepackage{mathtools} % for \DeclarePairedDelimiter
\usepackage{environ} % for \NewEnviron

\input{kl/config}
\input{header/config}
\input{header/macros}
\input{kl/abbreviations}
\input{kl/complexity}
\input{kl/paper}

\bibliographystyle{plainurl}% the mandatory bibstyle

\title{Tightening Faster Model Checking of First-Order Graph Properties, via Tarski's Calculus of Relations \begin{sideyoshiki}Tentative\end{sideyoshiki}} %TODO Please add
% On Faster Model Checking of First-Order Graph Properties and Tarski's Calculus of Relations

\titlerunning{FO3}

\author{Yoshiki Nakamura}{Science Tokyo, Japan}{mail}{https://orcid.org/0000-0003-4106-0408}{}
\author{Yuya Uezato}{CyberAgent, Inc., Japan}{mail}{https://orcid.org/0009-0005-8834-010X}{}

\authorrunning{Y. Nakamura and Y. Uezato}

\Copyright{Yoshiki Nakamura and Yuya Uezato}

\ccsdesc[500]{Theory of computation~Logic}

\keywords{First-order Logic,Fine-grained Complexity}

\category{} %optional, e.g. invited paper

\relatedversion{}

%\nolinenumbers %uncomment to disable line numbering

\begin{document}


\maketitle

\begin{abstract}
\input{sec/abstract}
\end{abstract}

\section{Notation}

\begin{hidden}
\Ja{
（メモ:
LIPICSとpLaTeXは異常に組み合わせが悪いので、
コメントなどで日本語を書く時はこのようにします。）}

\begin{screen}
Cute round-corner box \\
\Ja{コメントを付ける時には、screenで囲うとこういう感じになります}
\end{screen}

\begin{yoshiki}
  \Ja{コメントテスト}
\end{yoshiki}

\begin{sideyoshiki}
  \Ja{コメントテスト２}
\end{sideyoshiki}
\end{hidden}

We write $\fo^k$ to denote the set of formulas of First-order predicate logic with $k$-variables.

\paragraph*{Important Remark (Unary and Binary Predicates)}
We allow unary and binary predicates in our $\fo^3$.
Therefore, we do not allow terms involving ternary predicates such as $\forall x.\,\exists y.\,\forall z.\, P(x, y, z)$.%
\footnote{For each unary predicate $U$, we can simulate it by a diagonal binary predicate $B_U(x, y) := x = y \land U(x)$. However, in our later construction, we use some unary predicates; thus, we also allow unary predicates explicitly.}

On this remark, we can identify each model $\mathcal{M}$ as color graphs as follows:
\begin{itemize}
\item Each element corresponds a node (vertex).
\item Each unary predicate $U_i$ means ``color-$i$'' node.
\item Each binary predicate $P_j$ is ``color-$j$'' edge.
\item For example, on two nodes $x, y$, a term $(U_i(x) \land U_j(y) \land P_k(x, y))$ means that $x$ has the color $i$, $y$ has the color $j$, and there is a color $k$ edge from $x$ to $y$.
\end{itemize}

\paragraph*{Quantifier width}
\begin{Jcomment}
Quantifier widthという謎の指標を導入します。

与えられた式の部分項 $\forall/\exists x. ( \cdots )$ について $\cdots$部分に「直接」出てくるquantifierの個数を、この部分項のquantifier widthと呼ぶことにします。

たとえば、次の部分項
$$
\forall x.\biggl( \Bigl(\underline{{\color{red}\forall}} y. P(y)\Bigr) \lor \Bigl(\underline{{\color{red}\forall}} z. (\underline{\forall} x. P(z, x)) \land Q(z)\Bigr) \biggr)
$$
については、内部に3つのquantifierの出現（下線）があるのですが、直接の出現は赤くした2つだけなので、この部分項のquantifier widthは「２」です。

そして、式のquantifier widthというのは、最大のquantifier widthを持つ部分項における、その数として定義します。
\end{Jcomment}

\input{sec/introduction}
\input{sec/preliminaries}
\input{sec/fpt}
\input{sec/from_dominating_set}
\input{sec/from_cnf}
\input{sec/conclusion}

\bibliography{main.bib}
% \bibliography{../../../../library.bib} % for yn

\end{document}
